---
title: 'Introduction to R'
author: 'Caroline Hutchings'
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document: 
    code_folding: show
    theme: 'flatly'
    toc: true
    toc_float: true

---

## Vectors

- Create a vector of numbers using assignment operator
```{r message=FALSE, echo=FALSE, warning=FALSE, results=FALSE}
nums <- c(1,2,3) #assignment operator
nums
```
- Create a vector of strings
```{r}
week <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
```
- Create a vector with a range of numbers
```{r}
seq(1,10)
```
- Length of vector
```{r}
length(week)
```
## Working with Vectors

- Create a named vector
```{r}
temps <- c('Mon'=88, 'Tue'=93, 'Wed'=95, 'Thu'=91, 'Fri'=86)
temps
```
- Retrieve names of vector
```{r}
names(temps)
```
- Add values to a vector
```{r}
temps <- append(temps, c('Sa'=81, 'Su'=78))
temps
```
## Indexing

- Identify item at specific index
```{r}
week[1]
```
- All items except at specified index
```{r}
week[-2]
```
- All items in vector of indexes
```{r}
week[c(1,3,5)]
```


## Working with Vector Operators

- Relational Operators
```{r}
temps == 81
temps > 88
temps != 81
```

- Write code to shorten the names of temps 
```{r }
names(temps) <- c('M', 'T', 'W', 'Th', 'F')
temps
```

- Write code to select temperature of Friday
```{r}
temps['F']
temps[5]
```
## Dataframes

- Create a dataframe from vectors
```{r}
months <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
num_days <- c(31,28,31,30,31,30,31,31,30,31,30,31)
seasons <- c('W', 'W', 'Sp', 'Sp', 'Sp', 'Su', 'Su', 'Su', 'F', 'F', 'F', 'W')
calendar_df <- data.frame(Month = months, NumDays = num_days, Season=seasons)
calendar_df

```
## Dataframe Manipulations

- Add rows to a dataframe
```{r}
calendar_df <- rbind(calendar_df, list('NewJan', 31, 'Bonus'), list('NewFeb', 29, 'Bonus'))
calendar_df
```
- Add columns to a dataframe
```{r}
avg_temp <- c(52.3, 56.1, 62, 68.9, 75.9, 80.7, 81.9, 82.7, 78.5, 70.7, 61.1, 54.3, 53.4, 0)
calendar_df <- cbind(calendar_df, AvgTemp = avg_temp)
calendar_df
```
## Logical Operators

- Find Months with 30 days that are warmer than 70 degrees.
- Find Months that have a temperature lesss than 70 degrees or are summer months.
```{r}
calendar_df$AvgTemp >=70 & calendar_df$NumDays==30 #AND
calendar_df$AvgTemp <=70 | calendar_df$Season=='Su'
```
## Dataframe Application

- Find proportion of months with 31 days that are above 70 degrees.
- Find proportion of months that have less than 31 days or are above 70 degrees.
- Add a code chunck and then write code to answer the questions.

### Solution
```{r}
mean(calendar_df$AvgTemp >70 & calendar_df$NumDays==31)
mean(calendar_df$AvgTemp >70 | calendar_df$NumDays<31)
```


```{r}
library(tidyverse)
#convert a dataframe to a tibble
calendar_tibble <- as_tibble(calendar_df)
is_tibble(calendar_tibble)
is_tibble(calendar_df)
```


## Load Data Set

- Load data using Kaggle CLI

```{r}

library(tidyverse)

system("kaggle --version")

# Download dataset using Kaggle CLI
system("kaggle datasets download -d zeeshier/student-admission-records")

# Unzip dataset
unzip("student-admission-records.zip")

# Import into R
student_admission <- read_csv("student_admission_record_dirty.csv")

# Check if it's a tibble
is_tibble(student_admission)

```
## View Data in Spreadsheet Format

- View data in interactive window
```{r}
View(student_admission)
```

## Missing Data

- Check for missing data
```{r}
#count of rows that have missing name
sum(is.na(student_admission$Name))

#count of NA values for all columns
student_admission %>% 
  apply(MARGIN = 2, function(col) sum(is.na(col)))

#count of rows with one or more missing values
sum(!complete.cases(student_admission))

#drop rows with one or more missing values
student_admission_dropna <- student_admission %>% 
  filter(complete.cases(student_admission))
nrow(student_admission_dropna)
```
- Replace missing values with 'Unknown'
```{r}
#replace Name with 'Unknown'
student_admission_replace_na <- student_admission %>% 
  mutate(Name = replace_na(Name, 'Unknown'))

nrow(student_admission_replace_na)

sum(!complete.cases(student_admission_replace_na))


#replace Name and Admission Status with 'Unknown'
student_admission_replace_na_two <- student_admission %>% 
  mutate(across(c(Name, `Admission Status`),~ replace_na(.,'Unknown')))

nrow(student_admission_replace_na_two)

sum(!complete.cases(student_admission_replace_na_two))

```
## Student Admission Data Project 



## Manipulate Data

### Examine Count of Unique Values
```{r}
#lapply is list apply and apply is column/matrix apply
apply(X=student_admission, MARGIN = 2, FUN = unique) %>% #MARGIN=2 means column fun=function
  lapply(FUN = length) %>% str()
```
### Example: examine Admission Status
```{r}

unique(student_admission$`Admission Status`)

```
### Select and Rename Columns
```{r}

student_admission <- student_admission %>%
  select(Name, Age, Gender, `High School Percentage`, `Admission Test Score`, `Admission Status`, City)

names(student_admission) <- c("Name", "Age", "Gender", "HS_Percentage", "TestScore", "Admit", "City")
```
### Change Data Types 
```{r}

student_admission <- student_admission %>%
  mutate(
    Age = as.numeric(Age),
    HS_Percentage = as.numeric(HS_Percentage),
    TestScore = as.numeric(TestScore),
    Admit = ifelse(Admit == "Accepted", TRUE, FALSE)
  )
```
### Handling Missing Values

```{r}

student_admission <- student_admission %>%
  mutate(
    Name = replace_na(Name, "Unknown"),
    HS_Percentage = replace_na(HS_Percentage, mean(HS_Percentage, na.rm = TRUE)),
    TestScore = replace_na(TestScore, mean(TestScore, na.rm = TRUE))
  )

```


### Create Performance Category and Add to Tibble
```{r}

student_admission <- student_admission %>%
  mutate(
    Performance = case_when(
      HS_Percentage >= 85 ~ "High",
      HS_Percentage >= 70 ~ "Medium",
      TRUE ~ "Low"
    )
  )

View(student_admission)
```
### Create Combined Score
```{r}

student_admission <- student_admission %>%
  mutate(
    CombinedScore = HS_Percentage + TestScore
  )

View(student_admission)
```
### Group- Level Summary
```{r}

student_admission <- student_admission %>%
  group_by(City) %>%
  mutate(
    AvgCombinedScoreCity = mean(CombinedScore, na.rm = TRUE),
    AdmitRateCity = mean(Admit, na.rm = TRUE),
    CountCity = n()
  )
View(student_admission)
```
### City Summary

```{r}

city_summary <- student_admission %>%
  group_by(City) %>%
  summarize(
    MeanPercent = mean(HS_Percentage, na.rm = TRUE),
    MeanTest = mean(TestScore, na.rm = TRUE),
    AdmitRate = mean(Admit, na.rm = TRUE),
    Count = n()
  )

city_summary

```
### Admission Rate per Performance Group

```{r}

admit_pct <- student_admission %>%
  group_by(Performance) %>%
  summarize(
    Admitted = sum(Admit, na.rm = TRUE),
    Total = n()
  ) %>%
  mutate(
    AdmitPct = Admitted / Total
  )

admit_pct

```
### Graph Admission Rate by Performance

```{r}

ggplot(admit_pct, aes(x = Performance, y = AdmitPct, fill = Performance)) +
  geom_col() +
  labs(
    title = "Admission Rate by Performance Level",
    x = "Performance Category",
    y = "Admission Rate"
  ) +
  theme_minimal()

```
### Interactive Graph using Plotly

```{r}
#install.packages("plotly")
library(plotly)

p <- ggplot(admit_pct, aes(x = Performance, y = AdmitPct, fill = Performance)) +
  geom_col()

ggplotly(p)

```


